---
title: "Group membership models using lmerMultiMember"
output: rmarkdown::html_vignette
author:
    - Ben Bolker, adapted by Phillip Alday
vignette: >
  %\VignetteIndexEntry{Group membership models using lmerMultiMember}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r}
library(lmerMultiMember)
library(broom.mixed)
library(ggplot2); theme_set(theme_bw())
library(Matrix)
```

Consider a situation where we have multiple items of some sort which have associated effects, but we can’t association observations with single items; instead, each observation is associated with a group of items. (Two examples which have come up are (1) authorship of papers and (2) hockey players on a team.) These are not quite identical to "multi-membership models" (in part because group membership is binary, i.e., not weighted), although similar techniques to those shown here could work for multi-membership models.

Load packages (we don't really need anything beyond lmerMultiMember; the rest are for convenience/drawing pictures).

Construct a simulated example: first, simulate the design (structure).

```{r}
nm <- 20
nobs <- 500
set.seed(101)
## choose items for observations
pres <- matrix(rbinom(nobs*nm, prob=0.25, size=1),nrow=nobs,ncol=nm)
dimnames(pres) <- list(NULL,LETTERS[seq(nm)])
pres[1:5,]
```

```{r}
table(rowSums(pres))
```

```{r, fig.width=6, fig.height=4}
image(Matrix(pres),ylim=c(1,10),sub="",ylab="Observation",
      xlab="Item",
      ## draw tick labels at top
      scales=list(at=1:20,x=list(labels=colnames(pres),
                                 alternating=2)))
```

Since we chose which items/individuals to include for each observation randomly and independently (Bernoulli with probability 0.25), we have a highly variable number of individuals present for different observations (1-11). This would be realistic for some examples (authorship), unrealistic for others (hockey) ... I don't think it really makes much difference computationally (statistically, having some observations with a single member must make estimation more powerful ...) The 0/1 matrix (indicator variable for whether item $i$ is included in observation $j$) is convenient, and will turn out to be the form we need for inclusion in the model. It should be fairly straightforward to convert other forms (e.g. a list of sets of items associated with each observation) to this form ...

Now simulate the response variable.

```{r, results=FALSE}
b <- rnorm(nm)  ## item-level effects
## n.b. get in trouble if we don't add residual error
## (theta is scaled relative to residual error)
## here, use theta=sigma=1
y <- c(pres %*% b) +rnorm(nobs,sd=1)
```


Fit and the results look OK (correct item and residual variance estimated):
```{r}
m1 <- lmer(y ~ 1 + (1|membership), 
           data=data.frame(y=y, x=rep(1, nobs)), 
           memberships=list(membership=t(pres)))
```

The conditional modes by item look good:
```{r, fig.width=6, fig.height=4}
dd <- tidy(m1, effects="ran_vals")
dd <- transform(dd, level=reorder(level,estimate))
truth <- data.frame(level=LETTERS[seq(nm)],estimate=b)
ggplot(dd,aes(x=level,y=estimate))+
    geom_pointrange(aes(ymin=estimate-2*std.error,
              ymax=estimate+2*std.error))+coord_flip()+
    geom_point(data=truth,colour="red")
```

Session info:
```{r}
sessionInfo()
```

